Parametrization of Neural Networks with Connected Abelian Lie Groups as Data Manifold

### 2020-05-06 llncs2e/paper77.tex
corrections equation
typos

\tikzset{->-/.style={decoration={markings, mark=at position #1 with {\arrow[line width=1pt]{>}}},postaction={decorate}}}
\begin{document}
%
\title{Parametrization of Neural Networks with Connected Abelian Lie Groups as Data Manifold}
%
\titlerunning{Neural Networks on Connected Abelian Lie Groups}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Luciano Melodia\orcidID{0000-0002-7584-7287} \\
\and Richard Lenz\orcidID{0000-0003-1551-4824}}
%
\authorrunning{L. Melodia et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Chair of Computer Science 6\\
Friedrich-Alexander University Erlangen-Nürnberg\\
91058 Erlangen, Germany \\
\email{\{luciano.melodia,richard.lenz\}@fau.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Neural nets have been used in an elusive number of scientific disciplines. Nevertheless, their parameterization is largely unexplored. Dense nets are the coordinate transformations of a manifold from which the data is sampled. After processing through a layer, the representation of the original manifold may change. This is crucial for the preservation of its topological structure and should therefore be parameterized correctly. We discuss a method to determine the smallest topology preserving layer considering the data domain as abelian connected Lie group and observe that it is decomposable into $\mathbb{R}^p \times \mathbb {T}^q$. Persistent homology allows us to count its $k$-th homology groups. Using Künneth's theorem, we count the $k$-th Betti numbers. Since we know the embedding dimension of $\mathbb{R}^p$ and $\mathcal{S}^1$, we parameterize the bottleneck layer with the smallest possible matrix group, which can represent a manifold with those homology groups. Resnets guarantee smaller embeddings due to the dimension of their state space representation.

@@

We use invertible architectures, to demonstrate our hypothesis on a different Lie group from Euclidean space \citemain{NguyenAK19}. Our interpretation of a neural net is a continuous transformation, transferring invariant structures of the data in the input space to its deepest layer. Thus, it is capable to reconstruct the structure of the original space. This approach corresponds to the concept of homotopy equivalence \citemain{RubinsteinR08,YangY93}, which means that there is a continuous map $f: X \rightarrow Y$ and $g: Y \rightarrow X$, such that $f \circ g$ is homotopic to the identity $\text{Id}_Y$ and $g \circ f$ to $\text{Id}_X$. Thus, the spaces can be deformed continuously into one another. To do so, their homotopy groups must be equal. The fundamental group gives information about connected components which can be collapsed to a point by continuous functions, the first homotopy group about `holes' and the second about `cavities' and so forth. But homotopies are difficult to compute \citemain{klaus2004quick,hatcher2002algebraic}. Homology theory gives a remedy that guarantees an isomorphism between homotopy groups and homology groups under the conditions of Hurewicz's Theorem \citemain{klaus2004quick}. We use theoretical results from Lie theory and persistent homology theory to estimate how many dimensions of embedding are needed to represent the homology groups. Our paper is structured as follows:
\begin{itemize}
    \item In Sect. \ref{sec:manifoldassumption} we discuss the mathematical assumptions and recall the principles of smooth manifolds and Lie theory. In particular, we recall the decomposition of connected commutative Lie groups into $\mathbb{R}^p \times \mathbb{T}^q$.
    \item In Sect. \ref{sec:persistenthomology} we introduce to the theory of persistent homology, define persistence diagrams, persistence landscapes and confidence sets, in order to estimate homology groups despite noisy data. We use Künneth's theorem and derive how to count the homology groups of the input space.
    \item In Sect. \ref{sec:neuralnettheory} we explain the mathematical theory of neural nets and construct invertible deep neural nets to satisfy our assumptions. Resnets, which have a larger state space dimension, offer an architecture with a comparatively small number of parameters to dense nets.
    \item In Sect. \ref{sec:numericalexperiments} we validate our theory training autoencoders on the cifar10 and cifar100 datasets, isolating the innermost layer such that no residual connections bypass it on different embedding dimensions.
\end{itemize}

\section{The Manifold Assumption}
\label{sec:manifoldassumption}
The manifold assumption states that a point $x = \{x_1, \cdots, x_d\} \in \mathbb{R}^n$ lies on an embedded manifold $M$, with $\dim M \ll n$. We specify the concept of smooth manifolds. Then we dive into Lie groups concluding with a central theorem on the isomorphism of abelian connected Lie groups to the product space $\mathbb{R}^p \times \mathbb{T}^q$, which turns out useful to estimate the embedding dimension of the manifold.

\subsection{Background: Smooth Manifolds}
Manifolds generalize the concept of space mathematically. They are described by a collection of local coordinate maps $\varphi: U \rightarrow \varphi(U) \subseteq \mathbb{R}^n$ which are homeomorphic to $\mathbb{R}^{n}$ where $n = \text{dim} \; M$ and an open neighborhood $(U_i,\varphi_i)$ in $M$ which is covered by this map. The union of such maps should cover the entire space, i.e. $M = \bigcup_{i \in I} U_i$ and is called \textit{atlas} $\mathcal{A} = \{(U_i,\varphi_i) \; \vert \; i \in I\}$ \citemain{lee2013smooth}. The functions $x_1, \cdots, x_n: U \rightarrow \mathbb{R}$ are called local coordinates $\varphi(p) = (x_1(p), \cdots, x_n(p))$.

@@

A \textit{singular $k$-simplex} is defined as a continuous map $\sigma: \Delta^k \rightarrow X$ into a topological space $X$. Intuitively, every simplicial complex $K$ has some connected components. These form a vector space $C_k(X)$, which is called \textit{$k$-th chain group}, with coefficients in $\mathbb{Z}$ and elements called $k$-\textit{chains}. The $k$-chains are finite sums $\sum_{i=0}^{k} z_i \sigma_i$ with $z_i \in \mathbb{Z}$ \citemain[p.~108]{hatcher2002algebraic}. Having connected components defined as a group implies also the definition of a boundary. As the boundaries of a $k$-simplex are also $(k-1)$-simplices, we can remove all other simplices to project into the lower chain group by the $k$-th boundary operator $\partial_k: C_k(X) \rightarrow C_{k-1}(X)$:
\begin{equation}
    \partial_k(\sigma) = \sum_{i=0}^{k} (-1)^{i} \sigma \bigg\vert {[v_0,\cdots,\hat{v}_i, \cdots, v_k]}.
\end{equation}
Observe that the composition of boundary maps is zero, as a boundary has no boundary by itself, i.e. $\partial_i\partial_{i+1} = 0$ \citemain[pp.~108--109]{hatcher2002algebraic}. Thus, $\ker \; \partial_i$ corresponds to the generating elements of the $i$-th boundary group and $\Ima \; \partial_{i+1}$ for the $i$-th chain group. The singular homology group is the quotient space
\begin{equation}
    H_k(X) = \frac{\ker \left(\partial_k: C_k\left(X\right) \rightarrow C_{k-1}\left(X\right)\right)}{\Ima \left(\partial_{k+1}:C_{k+1}\left(X\right) \rightarrow C_{k}\left(X\right)\right)}.
\end{equation}
The $k$-th singular homology group is isomorphic to the $k$-th simplicial homology group for a trianguliable metric space, i.e. $H_k(X) \simeq H^{\Delta}_k(X)$ for all $k \in \mathbb{Z}$, see Appx. \ref{simplicialsingularequivalence}, thus we can transfer results from one homology theory to the other.

@@

\subsection{Persistence Diagrams}
Recording the \emph{Betti numbers} of the $k$-th homology group along the filtration, we obtain the \emph{$k$-dimensional persistence diagram}. The Betti numbers $\beta_k$ are defined by $\rank(H_k(X))$. We write $H^{\omega_i}_k = H_k(K^{\omega_i}(X))$ as $k$-th homology group on the simplicial complex $K$ with parametrization $\omega_i$. Then $H^{\omega_{i}}_{k} \rightarrow H^{\omega_{i+1}}_k$ induces a sequence of homomorphisms on the filtration, see Appx. \ref{persistenthomologyproofs} \citemain{lume,edelsbrunner2008persistent}:
\begin{align}
0 = H^{\omega_{0}}_k \rightarrow H^{\omega_{1}}_k \rightarrow \cdots \rightarrow H^{\omega_{n}}_k \rightarrow H^{\omega_{n+1}}_k = 0.
\end{align}
The image of each homomorphism consists of all $k$-dimensional homology classes which are born in $K^{\omega_i}(X)$ or appear before and die after spanning $K^{\omega_{i+1}}(X)$. Tracking the Betti numbers on the filtration results into a multiplicity
\begin{equation}
\mu^{\omega_{i},\omega_{j}}_k := (\beta_k^{\omega_{i},\omega_{j-1}} - \beta_k^{\omega_i,\omega_{j}})-(\beta_k^{\omega_{i-1},\omega_{j-1}}-\beta_k^{\omega_{i-1},\omega_{j}})
\end{equation}
for the $k$-th homology group and index pairs $(\omega_i,\omega_{j+1}) \in \overline{\mathbb{R}^2} := \mathbb{R}^2 \cup \infty$ with indices $i \leq j$ \citemain{edelsbrunner2008persistent}. The Euclidean space is extended, as the first connected component of our filtration is not merged with any other critical value. We assign the value $\infty$ to it, corresponding to the second coordinate $w_{j+1}$ in the persistence diagram. The first term counts elements born in $K^{\omega_{j-1}}(X)$ and vanished entering $K^{\omega_{j}}(X)$, while the second term counts the representatives of homology classes before $K^{\omega_{j}}(X)$ and vanishing at $K^{\omega_{j}}(X)$. The index pairs denote the \emph{birth} and \emph{death} of a homology class. The $k$-th persistence diagram is thus
\begin{equation}
\mathfrak{P}_k(K(X)) := \left\{(\omega_i, \omega_{j+1}) \in \overline{\mathbb{R}^2} \; \bigg\vert \; \mu^{\omega_{i},\omega_{j+1}}_k = 1 \; \text{for all} \; i,j \in I \right\}.
\end{equation}

@@

\subsection{Counting Betti Numbers}
In the following, the homology groups are computed over $\mathbb{Z}$. Based on our assumption to operate on a connected commutative Lie group, there is an isomorphism for the homology groups of the data space $G$, so that $H_k(G) \simeq H_k(\mathbb{R}^p \times \mathcal{S}^{1}_{1}\times\cdots\times\mathcal{S}^{1}_{q})$ with $p+q = \dim G$, see \ref{decomposition}. The homology groups of this product space are equal to the homology groups of the sought manifold. The neural net should be able to preserve all structures within this space. Since the input domain is a set of points to be described by one or more compact manifolds, all homology classes should be preserved in our learning task. Künneth's theorem states that the $k$-th singular homology group of a topological product space $X \times Y$ is isomorphic to the product of the respective $k$-th homology groups of $X$ and $Y$, see \citemain[p.~268]{hatcher2002algebraic} and Appx. \ref{a:groups}:
\begin{align}
    H_k(X \times Y) \simeq \bigoplus_{i+j=k}H_i(X) \otimes H_j(Y).
\end{align}
This applies to all fields, since modules over a field are always free. If we apply the theorem to the decomposition, we get for the space we are looking for
\begin{align}
	\label{homologykunneth}
    &H_k(\mathcal{S}^{1}_{1}\times\cdots\times \mathcal{S}^{1}_{q}\times\mathbb{R}^p) \simeq \\
    \simeq \bigoplus_{i_{1} + \cdots + i_{r}=k} &H_{i_{1}}(\mathcal{S}^{1}_{1}) \otimes \cdots \otimes H_{i_{\hat{r}}}(\mathcal{S}^{1}_{q}) \otimes H_{i_{r}}(\mathbb{R}^{p}).
\end{align}
We need to know how to count in order to derive the dimension of the sought manifold from the representatives of homology classes within the mean persistence landscapes. Note, in Eq. \ref{homologykunneth} terms remain only for indices $i_{j} \in \{0,1\}$. Thus, the homology group $H_k(\mathbb{T}^{q}) \simeq H_k(\mathcal{S}^{1}_{1} \times \cdots \times \mathcal{S}^{1}_{q}) \simeq \mathbb{Z}^{{q\choose{k}}}$ and $H_0(\mathbb{R}^p)\simeq\mathbb{Z}^{p}$, where $p$ indicates the number of connected components. We denote the amount of representatives from homology classes of the $k$-th homology group appearing in the mean persistence landscape $\bar{\mathfrak{L}}$ by $\sharp^{H}_{k}(\bar{\mathfrak{L}}(\mathfrak{P}(K(X)))$. Thus, we get the following relation for the representatives of the zeroth homology group:
\begin{align}
	\label{countingsimple}
	\sharp_{0}^{H}(\bar{\mathfrak{L}}(K(X))) \approx p.
\end{align}
This is the number of connected components to be represented by the innermost layer of the neural net, corresponding exactly to the number of elements from the zeroth homology group. We only need to consider the homology groups which are represented in the mean persistence landscape. Assuming that our estimate is good, higher homology groups are trivial. The required number of factors of $1$-spheres can be determined for the $k$-th homology group by
\begin{align}
	\label{countingequation}
	\sharp_{k}^{H}(\bar{\mathfrak{L}}(\mathfrak{P}(K(X))) \approx \prod_{i=1}^{k} \frac{q-k+i}{i}.
\end{align}
Note, that we assume to work on a smooth manifold, whose de Rham cohomology induces an isomorphism into the simplicial homology with coefficients in $\mathbb{R}$, see Appx. \ref{simplicialsingularequivalence}. We can still work with coefficients in $\mathbb{Z}$ because, according to the Universal Coefficient Theorem, the integral homology groups determine the homology groups over every field \cite{gruenberg1968universal}. Thus, these results can be applied without restriction to any homology theory, see Appx. \ref{simplicialsingularequivalence}.

@@

$\sigma$ denotes a non-linear activation function. Commonly used are functions from the exponential family. A feedforward net is then a map $\varphi^{(l)}:\textbf{x}^{(l)}(M)$ $\rightarrow$ $\left(\varphi^{(l)}\circ \textbf{x}^{(l)}\right)(M)$. $\textbf{x}^{(0)} \in M$ are points on a Riemann manifold $M$. Learning by back propagation can be considered as a change of coordinate representation or shift on the tangent bundle of the underlying manifold \citemain{rumelhart1986learning}.

\subsection{Deep Learning}
\input{figure2.tex}
Neural nets are used in practice with a different number of perceptrons per layer. However, a layer, from the perspective of a matrix group, can also represent the manifold of the previous one. Manifolds can always be immersed and submersed as long as the rank of the Jacobian of the respective map remains constant. Thus, the dimension of the data manifold is the dimension of the smallest possible bottleneck layer of the neural net  $\dim M = \min_l \dim \textbf{x}^{(l)}$ \citemain{HauserR17}. Our contribution provides the theory of how this dimension can be determined for all neural nets acting on an abelian connected Lie group as domain of the data. The concatenation of such maps is called a deep neural net:
\begin{equation}
    \textbf{x}^{(L)}(M) = \left(\varphi^{(l+1)} \circ \varphi^{(l)} \cdots \circ \varphi^{(1)} \circ \varphi^{(0)} \circ \textbf{x}^{(0)}\right)(M).
\end{equation}
Activation functions restrict to a subgroup of Euclidean space. Continuous homomorphisms between Lie groups are automatically smooth. Thus, a second countable locally Euclidean group has exactly one differentiable structure, turning it into a Lie group. We need to show that $\textbf{x}^{(l+1)}$ has the topology of $G$ and is therefore an embedded closed Lie subgroup $\textbf{x}^{(l)} := H \subseteq G$. Cartans Theorem of Closed Subgroups states that a subgroup $H$ of a Lie group $G$ is an embedded Lie subgroup, if it is a closed subgroup \citemain{cartan1952theorie}. We construct our neural nets such that they are a subset of $\text{GL}(n,\mathbb{R})$ and according to the proof in Appx. \ref{liegroups}, this is a closed subgroup of $\text{GL}(n,\mathbb{C})$, which is itself a Lie group. It follows that a layer in the neural net is also a Lie group. Due to invertibility, the subsequent smaller layers are Lie subgroups, as $\sigma^{(l)}$ is smooth.

@@

\paragraph{Acknowledgements} The work is partially supported by Siemens Power and Gas. The authors would like to thank Noah Becker and David Haller for proofreading.

\paragraph{Code} The code can be found at: \href{https://github.com/karhunenloeve/Irhe}{https://github.com/karhunenloeve/Irhe}.
%
%
%
%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystylemain{splncs04}
\bibliographymain{biblio}
%
\newpage
\appendix
\section{Supplementary Material for Sect. \ref{sec:manifoldassumption}}
\label{a:mannilliegroup}
\subsection{Groups}
\label{a:groups}
\begin{definition}
A set $G$ together with a map $\mu: G \times G \rightarrow G$, $(x,y) \mapsto xy$ and an element $e = e_G$ is called group if the following conditions are fulfilled:
\begin{enumerate}
	\item[(1)] $(xy)z = x(yz)$ for all $x,y,z \in G$, thus the elements are associative,
	\item[(2)] $xe = ex = x$ for all $x \in G$, so there is a neutral element and
	\item[(3)] $\forall x \in G \; \exists \; x^{-1} \in G$, such that $xx^{-1} = x^{-1}x = e$. 
\end{enumerate}
\end{definition}

>>>>

\tikzset{->-/.style={decoration={markings, mark=at position #1 with {\arrow[line width=1pt]{>}}},postaction={decorate}}}
\begin{document}
%
\title{Parametrization of Neural Networks with Connected Abelian Lie Groups as Data Manifold}
%
\titlerunning{Neural Networks on Connected Abelian Lie Groups}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Luciano Melodia\orcidID{0000-0002-7584-7287} \\
\and Richard Lenz\orcidID{0000-0003-1551-4824}}
%
\authorrunning{L. Melodia et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Chair of Computer Science 6\\
Friedrich-Alexander University Erlangen-Nürnberg\\
91058 Erlangen, Germany \\
\email{\{luciano.melodia,richard.lenz\}@fau.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Neural nets have been used in an substantial number of scientific disciplines. Despite this, their parameterization is largely unexplored. Dense nets are the coordinate transformations of a manifold from which the data is sampled. After a layer processes the data, the representation of the original manifold may change. This is crucial for the preservation of its topological structure and should therefore be parameterized correctly. We discuss a method to determine the smallest topology preserving layer considering the data domain as abelian connected Lie group and observe that it is decomposable into $\mathbb{R}^p \times \mathbb {T}^q$. Persistent homology allows us to count its $k$-th homology groups. Using Künneth's theorem, we count the $k$-th Betti numbers. Since we know the embedding dimension of $\mathbb{R}^p$ and $\mathcal{S}^1$, we parameterize the bottleneck layer with the smallest possible matrix group, which can represent a manifold with those homology groups. Resnets guarantee smaller embeddings due to the dimension of their state space representation.

@@

In the following we use invertible architectures, to demonstrate our hypothesis on a different Lie group from Euclidean space \citemain{NguyenAK19}. Our interpretation of a neural net is a continuous transformation, transferring invariant structures of the data in the input space to its deepest layer. Thus, it is able to reconstruct the structure of the original space. This approach corresponds to the concept of homotopy equivalence \citemain{RubinsteinR08,YangY93}, which means that there is a continuous map $f: X \rightarrow Y$ and $g: Y \rightarrow X$, such that $f \circ g$ is homotopic to the identity $\text{Id}_Y$ and $g \circ f$ to $\text{Id}_X$. Thus, the spaces can be deformed continuously into one another. To do so, their homotopy groups must be equal. The fundamental group gives information about connected components. These can be collapsed to a point by continuous functions, the first homotopy group about `holes' and the second about `cavities' and so forth. But homotopies are difficult to compute \citemain{klaus2004quick,hatcher2002algebraic}. Homology theory gives a remedy that guarantees an isomorphism between homotopy groups and homology groups under the conditions of Hurewicz's Theorem \citemain{klaus2004quick}. We use theoretical results from Lie theory and persistent homology theory to estimate how many dimensions of embedding are needed to represent the homology groups. Our paper is structured as follows:
\begin{itemize}
    \item In Sect. \ref{sec:manifoldassumption} we discuss the mathematical assumptions and recall the principles of smooth manifolds and Lie theory. In particular, we recall the decomposition of connected commutative Lie groups into $\mathbb{R}^p \times \mathbb{T}^q$.
    \item In Sect. \ref{sec:persistenthomology} we introduce the theory of persistent homology, define persistence diagrams, persistence landscapes and confidence sets, in order to estimate homology groups despite noisy data. We use Künneth's theorem and derive how to count the homology groups of the input space.
    \item In Sect. \ref{sec:neuralnettheory} we explain the mathematical theory of neural nets and construct invertible deep neural nets to satisfy our assumptions. Resnets, which have a larger state space dimension, offer an architecture with a comparatively small number of parameters to dense nets.
    \item In Sect. \ref{sec:numericalexperiments} we validate our theory training autoencoders on the cifar10 and cifar100 datasets, isolating the innermost layer such that no residual connections bypass it on different embedding dimensions.
\end{itemize}

\section{The Manifold Assumption}
\label{sec:manifoldassumption}
The manifold assumption states that a point $x = \{x_1, \cdots, x_d\} \in \mathbb{R}^n$ lies on an embedded manifold $M$, with $\dim M \ll n$. First, we specify the concept of smooth manifolds. Then we dive into Lie groups concluding with a central theorem on the isomorphism of abelian connected Lie groups to the product space $\mathbb{R}^p \times \mathbb{T}^q$, which turns out useful to estimate the embedding dimension of the manifold.

\subsection{Background: Smooth Manifolds}
Manifolds generalize the concept of space mathematically. They are described by a collection of local coordinate maps $\varphi: U \rightarrow \varphi(U) \subseteq \mathbb{R}^n$ which are homeomorphic to $\mathbb{R}^{n}$, where $n = \text{dim} \; M$, and an open neighborhood $(U_i,\varphi_i)$ in $M$ which is covered by this map. The union of such maps should cover the entire space, i.e. $M = \bigcup_{i \in I} U_i$ and is called \textit{atlas} $\mathcal{A} = \{(U_i,\varphi_i) \; \vert \; i \in I\}$ \citemain{lee2013smooth}. The functions $x_1, \cdots, x_n: U \rightarrow \mathbb{R}$ are called local coordinates $\varphi(p) = (x_1(p), \cdots, x_n(p))$.

@@

A \textit{singular $k$-simplex} is defined as a continuous map $\sigma: \Delta^k \rightarrow X$ into a topological space $X$. Intuitively, every simplicial complex $K$ has some connected components. These form a vector space $C_k(X)$, which is called \textit{$k$-th chain group}, with coefficients in $\mathbb{Z}$ and elements called $k$-\textit{chains}. The $k$-chains are finite sums $\sum_{i=0}^{k} z_i \sigma_i$ with $z_i \in \mathbb{Z}$ \citemain[p.~108]{hatcher2002algebraic}. Having connected components defined as a group also implies the definition of a boundary. As the boundaries of a $k$-simplex are also $(k-1)$-simplices, we can remove all other simplices to project into the lower chain group by the $k$-th boundary operator $\partial_k: C_k(X) \rightarrow C_{k-1}(X)$:
\begin{equation}
    \partial_k(\sigma) = \sum_{i=0}^{k} (-1)^{i} \sigma \bigg\vert {[v_0,\cdots,\hat{v}_i, \cdots, v_k]}.
\end{equation}
Observe that the composition of boundary maps is zero, as a boundary has no boundary by itself, i.e. $\partial_i\partial_{i+1} = 0$ \citemain[pp.~108--109]{hatcher2002algebraic}. Thus, $\ker \; \partial_i$ corresponds to the generating elements of the $i$-th boundary group and $\Ima \; \partial_{i+1}$ for the $i$-th chain group. The singular homology group is the quotient space
\begin{equation}
    H_k(X) = \frac{\ker \left(\partial_k: C_k\left(X\right) \rightarrow C_{k-1}\left(X\right)\right)}{\Ima \left(\partial_{k+1}:C_{k+1}\left(X\right) \rightarrow C_{k}\left(X\right)\right)}.
\end{equation}
The $k$-th singular homology group is isomorphic to the $k$-th simplicial homology group for a trianguliable metric space, i.e. $H_k(X) \simeq H^{\Delta}_k(X)$ for all $k \in \mathbb{Z}$, see Appx. \ref{simplicialsingularequivalence}, thus we can transfer results from one homology theory to the other.

@@

\subsection{Persistence Diagrams}
Recording the \emph{Betti numbers} of the $k$-th homology group along the filtration, we obtain the \emph{$k$-dimensional persistence diagram}. The Betti numbers $\beta_k$ are defined by $\rank(H_k(X))$. We write $H^{\omega_i}_k = H_k(K^{\omega_i}(X))$ as $k$-th homology group on the simplicial complex $K$ with parametrization $\omega_i$. Then $H^{\omega_{i}}_{k} \rightarrow H^{\omega_{i+1}}_k$ induces a sequence of homomorphisms on the filtration, see Appx. \ref{persistenthomologyproofs} \citemain{lume,edelsbrunner2008persistent}:
\begin{align}
0 = H^{\omega_{0}}_k \rightarrow H^{\omega_{1}}_k \rightarrow \cdots \rightarrow H^{\omega_{n}}_k \rightarrow H^{\omega_{n+1}}_k = 0.
\end{align}
The image of each homomorphism consists of all $k$-dimensional homology classes which are born in $K^{\omega_i}(X)$ or appear before and die after spanning $K^{\omega_{i+1}}(X)$. Tracking the Betti numbers on the filtration results into a multiplicity
\begin{equation}
\mu^{\omega_{i},\omega_{j}}_k := (\beta_k^{\omega_{i},\omega_{j-1}} - \beta_k^{\omega_i,\omega_{j}})-(\beta_k^{\omega_{i-1},\omega_{j-1}}-\beta_k^{\omega_{i-1},\omega_{j}})
\end{equation}
for the $k$-th homology group and index pairs $(\omega_i,\omega_{j+1}) \in \overline{\mathbb{R}^2} := \mathbb{R}^2 \cup \infty$ with indices $i \leq j$ \citemain{edelsbrunner2008persistent}. The Euclidean space is extended, as the first connected component of our filtration is not merged with any other critical value. We assign the value $\infty$ to it, corresponding to the second coordinate $w_{j+1}$ in the persistence diagram. The first term counts elements born in $K^{\omega_{j-1}}(X)$ and which vanish entering $K^{\omega_{j}}(X)$, while the second term counts the representatives of homology classes before $K^{\omega_{j}}(X)$ and which vanish at $K^{\omega_{j}}(X)$. The index pairs denote the \emph{birth} and \emph{death} of a homology class. The $k$-th persistence diagram is thus
\begin{equation}
\mathfrak{P}_k(K(X)) := \left\{(\omega_i, \omega_{j+1}) \in \overline{\mathbb{R}^2} \; \bigg\vert \; \mu^{\omega_{i},\omega_{j+1}}_k = 1 \; \text{for all} \; i,j \in I \right\}.
\end{equation}

@@

\subsection{Counting Betti Numbers}
In the following, the homology groups are computed over $\mathbb{Z}$. Based on our assumption to operate on a connected commutative Lie group, there is an isomorphism for the homology groups of the data space $G$, so that $H_k(G) \simeq H_k(\mathbb{R}^p \times \mathcal{S}^{1}_{1}\times\cdots\times\mathcal{S}^{1}_{q})$ with $p+q = \dim G$, see \ref{decomposition}. The homology groups of this product space are equal to the homology groups of the sought manifold. The neural net should be able to preserve all structures within this space. Since the input domain is a set of points to be described by one or more compact manifolds, all homology classes should be preserved in our learning task. Künneth's theorem states that the $k$-th singular homology group of a topological product space $X \times Y$ is isomorphic to the product of the respective $k$-th homology groups of $X$ and $Y$, see \citemain[p.~268]{hatcher2002algebraic} and Appx. \ref{a:groups}:
\begin{align}
    H_k(X \times Y) \simeq \bigoplus_{i+j=k}H_i(X) \otimes H_j(Y).
\end{align}
This applies to all fields, since modules over a field are always free. If we apply the theorem to the decomposition, we get for the space we are looking for
\begin{align}
	\label{homologykunneth}
    &H_k(\mathcal{S}^{1}_{1}\times\cdots\times \mathcal{S}^{1}_{q}\times\mathbb{R}^p) \simeq \\
    \simeq \bigoplus_{i_{1} + \cdots + i_{r}=k} &H_{i_{1}}(\mathcal{S}^{1}_{1}) \otimes \cdots \otimes H_{i_{\hat{r}}}(\mathcal{S}^{1}_{q}) \otimes H_{i_{r}}(\mathbb{R}^{p}).
\end{align}
We need to count the representatives of homology classes within the mean persistence landscapes in order to derive the dimension of the sought manifold. Note, in Eq. \ref{homologykunneth} terms remain only for indices $i_{j} \in \{0,1\}$. Thus, the homology group $H_k(\mathbb{T}^{q}) \simeq H_k(\mathcal{S}^{1}_{1} \times \cdots \times \mathcal{S}^{1}_{q}) \simeq \mathbb{Z}^{{q\choose{k}}}$ and $H_0(\mathbb{R}^p)\simeq\mathbb{Z}^{p}$, where $p$ indicates the number of connected components. We denote the amount of representatives from homology classes of the $k$-th homology group appearing in the mean persistence landscape $\bar{\mathfrak{L}}$ by $\sharp^{H}_{k}(\bar{\mathfrak{L}}(\mathfrak{P}(K(X)))$. Thus, we get the following relation for the representatives of the zeroth homology group:
\begin{align}
	\label{countingsimple}
	\sharp_{0}^{H}(\bar{\mathfrak{L}}(\mathfrak{P}(K(X)))) \approx p.
\end{align}
This is the number of connected components to be represented by the innermost layer of the neural net, corresponding exactly to the number of elements from the zeroth homology group. We only need to consider the homology groups which are represented in the mean persistence landscape. Assuming that our estimate is good, higher homology groups are trivial. The required number of factors of $1$-spheres can be determined for the $k$-th homology group by
\begin{align}
	\label{countingequation}
	\sharp_{k}^{H}(\bar{\mathfrak{L}}(\mathfrak{P}(K(X))) \approx \prod_{i=1}^{k} \frac{q-k+i}{i}.
\end{align}
Note, that we assume to work on a smooth manifold, whose de Rham cohomology induces an isomorphism into the simplicial homology with coefficients in $\mathbb{R}$, see Appx. \ref{simplicialsingularequivalence}. We can still work with coefficients in $\mathbb{Z}$ because, according to the Universal Coefficient Theorem, the integral homology groups determine the homology groups over every field \cite{gruenberg1968universal}. Thus, these results can be applied without restriction to any homology theory, see Appx. \ref{simplicialsingularequivalence}.

@@

$\sigma$ denotes a non-linear activation function. Commonly used are functions from the exponential family. As a result a feedforward net is then a map $\varphi^{(l)}:\textbf{x}^{(l)}(M)$ $\rightarrow$ $\left(\varphi^{(l)}\circ \textbf{x}^{(l)}\right)(M)$. $\textbf{x}^{(0)} \in M$ are points on a Riemann manifold $M$. Learning by back propagation can be considered as a change of coordinate representation or shift on the tangent bundle of the underlying manifold \citemain{rumelhart1986learning}.

\subsection{Deep Learning}
\input{figure2.tex}
In practice neural nets are used with a different number of perceptrons per layer. However, a layer, from the perspective of a matrix group, can also represent the manifold of the previous one. Manifolds can always be immersed and submersed as long as the rank of the Jacobian of the respective map remains constant. Thus, the dimension of the data manifold is the dimension of the smallest possible bottleneck layer of the neural net  $\dim M = \min_l \dim \textbf{x}^{(l)}$ \citemain{HauserR17}. Our contribution provides the theory of how this dimension can be determined for all neural nets acting on an abelian connected Lie group as domain of the data. The concatenation of such maps is called a deep neural net:
\begin{equation}
    \textbf{x}^{(L)}(M) = \left(\varphi^{(l+1)} \circ \varphi^{(l)} \cdots \circ \varphi^{(1)} \circ \varphi^{(0)} \circ \textbf{x}^{(0)}\right)(M).
\end{equation}
Activation functions restrict to a subgroup of Euclidean space. Continuous homomorphisms between Lie groups are automatically smooth. Thus, a second countable locally Euclidean group has exactly one differentiable structure, turning it into a Lie group. We need to show that $\textbf{x}^{(l+1)}$ has the topology of $G$ and is therefore an embedded closed Lie subgroup $\textbf{x}^{(l)} := H \subseteq G$. Cartans Theorem of Closed Subgroups states that a subgroup $H$ of a Lie group $G$ is an embedded Lie subgroup, if it is a closed subgroup \citemain{cartan1952theorie}. We construct our neural nets such that they are a subset of $\text{GL}(n,\mathbb{R})$ and according to the proof in Appx. \ref{liegroups}, this is a closed subgroup of $\text{GL}(n,\mathbb{C})$, which is itself a Lie group. It follows that a layer in the neural net is also a Lie group. Due to invertibility, the subsequent smaller layers are Lie subgroups, as $\sigma^{(l)}$ is smooth.

@@

\subsubsection*{Acknowledgements} The work is partially supported by Siemens Power and Gas. The authors thank Noah Becker and David Haller for proofreading.

\subsubsection*{Code} The code can be found at: \href{https://github.com/karhunenloeve/Irhe}{https://github.com/karhunenloeve/Irhe}.
%
%
%
%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystylemain{splncs04}
\bibliographymain{biblio}
%
\newpage
\appendix
\section{Supplementary Material for Sect. \ref{sec:manifoldassumption}}
\label{a:mannilliegroup}
\subsection{Groups}
\label{a:groups}
\begin{definition}
A set $G$ together with a map $\mu: G \times G \rightarrow G$, $(x,y) \mapsto xy$ and an element $e = e_G$ is called group if the following conditions are fulfilled:
\begin{enumerate}
	\item[(1)] $(xy)z = x(yz)$ for all $x,y,z \in G$, thus the elements are associative,
	\item[(2)] $xe = ex = x$ for all $x \in G$, so there is a neutral element and
	\item[(3)] $\forall x \in G \; \exists \; x^{-1} \in G$, such that $xx^{-1} = x^{-1}x = e$. 
\end{enumerate}
\end{definition}
